Mamba vs SOTA for Vision Tasks\\ \thanks{$\dagger$ Equal contributions}

### 2024-04-30 paper/main.tex
paper update

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{algorithm}

@@

\begin{abstract}
    Abstract
    This paper explores the performance of the Mamba architecture, a variant of Structured State Space Sequence Models (SSMs), in comparison to state-of-the-art models such as Transformers, U-Net, and ResNet for vision tasks. We focus on image segmentation and classification, with the goal of maintaining high accuracy while keeping parameter counts low. Developing compact models with fewer parameters is crucial for the future of AI, not only to reduce energy consumption but also to enable deployment on resource-constrained edge devices.
    By leveraging the selective state spaces in Mamba blocks, we aim to achieve efficient and effective segmentation and classification performance while maintaining the linear scalability of SSMs. We implement Mamba-based architectures and compare their results to popular models like U-Net, ResNet, and Transformer-based approaches on the tasks of segmentation and classification.
    In addition to accuracy, we place a strong emphasis on model size, targeting compact architectures suitable for resource-constrained environments. Through careful design choices and optimizations, we strive to develop Mamba-based models that achieve competitive accuracy with significantly fewer parameters compared to existing state-of-the-art models.
    This work highlights the potential of Mamba and SSMs for efficient vision tasks, contributing to the development of compact yet accurate models in image segmentation and classification. We open-source our code to facilitate further research and exploration of these architectures: \href{https://github.com/lxaw/mamba-vs-else-vision}{https://github.com/lxaw/mamba-vs-else-vision}
\end{abstract}

\begin{IEEEkeywords}
    Mamba, TinyML, Image Segmentation, Image Classification
\end{IEEEkeywords}

\section{Introduction}
The rapid advancement of deep learning has led to remarkable achievements in various domains, including computer vision, natural language processing, and speech recognition. However, this progress has been accompanied by a significant increase in the size and complexity of neural network models. State-of-the-art (SOTA) models often have several million or even billions of parameters, making them computationally expensive and challenging to deploy on resource-constrained devices \cite{brown2020language, dosovitskiy2021image}. This trend poses concerns regarding energy consumption and the feasibility of deploying these models on edge devices, which have limited memory and processing power.

@@

CNNs achieve their efficiency and effectiveness through their unique architecture and the use of convolutional operations. The convolutional layers in CNNs apply a set of learnable filters (kernels) to the input image or feature map. These filters slide over the input, performing element-wise multiplications and summing the results to produce a new feature map. This process allows CNNs to automatically learn and extract relevant features from the input data.

@@

@@

Now, the kernel can be constructed from the convolutional form \eqref{eq:convolutionalform} and output $y$ can be computed.
\begin{align}
    \mbfb{K} & = (\mathbf{C}\mbfb{B}, \mathbf{C}\mbfb{A}\mbfb{B}, \ldots, \mathbf{C}\mbfb{A}^k\mbfb{B}, \ldots) \\
    y        & = x * \mbfb{K}
\end{align}
This convolution can be computed in parallel as $y_k$ does not depend on previous outputs.

@@

For this our analysis, we compare four models against our own developed ones. The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{DBLP:journals/corr/HeZRS15}, ResNet50, U-Mamba\cite{U-Mamba}, and SegViT\cite{dosovitskiy2021image}. Below we explain our choices.
\subsection{U-Net}
The core idea behind U-Net is to complement a traditional contracting network with successive layers that replace pooling operations with upsampling operators. These upsampling layers aim to increase the resolution of the output. Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information \cite{ronneberger2015unet}.
A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part. This allows the network to effectively propagate contextual information to higher resolution layers. As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture. To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image. This tiling strategy is crucial for applying the network to large images, as it circumvents resolution limitations imposed by GPU memory constraints \cite{ronneberger2015unet}. Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models. For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{DBLP:journals/corr/abs-2006-11239} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}. We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{DBLP:journals/corr/HeZRS15}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegFormer}
SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA). The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks. The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution. This allows SegFormer to efficiently process high-resolution images while capturing both local and global context. Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details. By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

@@

\section{Results}
We now compare our models against ResNet18, ResNet50, standard U-Mamba bottleneck, UNet, and UltraLight VM-UNet. We train for 100 epochs on all models, and use dice-loss with Adam optimizer. We train on roughly 3000 images, and validate on roughly 400. we then test on roughly 400. We show the distributions of our data in Figure \ref{fig:datastats}.
Model results are shown in Table \ref{tab:model_comparison} and Figures \ref{fig:balls},\ref{fig:segs}.

\section{Results}
We now compare our models against ResNet18, ResNet50, standard U-Mamba bottleneck, UNet, and UltraLight VM-UNet. We train for 100 epochs on all models, and use dice-loss with Adam optimizer. We train on roughly 3000 images, and validate on roughly 400. we then test on roughly 400. We show the stats of our data in Figure \ref{fig:datastats}.
The results are shown in Figures \ref{fig:balls} and \ref{fig:segs}.

\begin{table}[ht]
    \centering
    \begin{tabular}{|l|c|c|}
        \hline
        \textbf{Model Name} & \textbf{Parameter Count (Millions)} & \textbf{Dice Score} \\
        \hline
        UMambaBot\_PP       & 10.0                                & 0.8867              \\
        UMambaBot           & 9.8                                 & 0.8799              \\
        ResNet18            & 13.9                                & 0.8648              \\
        ResNet50            & 42.9                                & 0.8672              \\
        SegFormer           & 17.8                                & 0.8454              \\
        UNet                & 7.8                                 & 0.8929              \\
        UL-VM-UNet          & 0.049                               & 0.79955             \\
        UL-VM-UNet\_v1      & 0.20                                & 0.8409              \\
        UL-VM-UNet\_v2      & 0.042                               & 0.8243              \\
        UL-VM-UNet\_v3      & 0.176                               & 0.8109              \\
        UL-VM-UNet\_v4      & 1.1                                 & 0.8548              \\
        UL-VM-UNet\_v5      & 0.7                                 & 0.8633              \\
        \hline
    \end{tabular}
    \caption{Comparison of different segmentation models.}
    \label{tab:model_comparison}
\end{table}

@@

To further demonstrate the viability and versatility of the Mamba architecture, we conducted experiments on the CIFAR-10 dataset \cite{krizhevsky2009learning}, a widely-used benchmark for image classification tasks. CIFAR-10 consists of 60,000 32x32 color images across 10 classes, with 6,000 images per class. The dataset is split into 50,000 training images and 10,000 test images.
For this experiment, we employed the Vision Mamba (ViM) architecture proposed by Zhu et al. \cite{zhu2024vision}. ViM is an efficient visual representation learning framework that leverages a bidirectional state space model to capture long-range dependencies in visual data. By combining the strengths of CNNs and SSMs, ViM aims to achieve competitive performance while maintaining computational efficiency.

@@

We compared the performance of ViM against the Vision Transformer (ViT) architecture \cite{dosovitskiy2021image} on the CIFAR-10 dataset. The ViM model had 1,974,664 parameters, while the ViT had 1,767,602.
Both models were trained for 100 epochs using the same experimental setup, including hyperparameters, optimization algorithm, and data augmentation techniques. The experiments were conducted on an NVIDIA GeForce RTX 3090 Ti GPU to ensure a fair comparison of training and inference times.
Figure \ref{fig:ViMViTF1} presents the validation F1 score and loss curves for both ViM and ViT over the course of 100 epochs, Figure \ref{fig:ViMViTLoss} shows the validation loss values over the 100 epochs, while Figure \ref{fig:ViMViTTrainingValidationTime} compares the training and validation times for both models. Note that ViM achieves similar performance with significantly less training / validation time.

@@

\section{Conclusion}

\begin{thebibliography}{00}
    \bibitem{brown2020language} T. B. Brown et al., "Language Models are Few-Shot Learners," arXiv:2005.14165 [cs.CL], 2020.
    \bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
    \bibitem{banbury2021benchmarking} C. R. Banbury et al., "MLPerf Tiny Benchmark," arXiv:2106.07597 [cs.LG], 2021.
    \bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.
    \bibitem{he2016deep} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," arXiv:1512.03385 [cs.CV], 2016.
    \bibitem{vaswani2017attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2017.
    \bibitem{choromanski2020rethinking} K. Choromanski et al., "Rethinking Attention with Performers," arXiv:2009.14794 [cs.LG], 2020.
    \bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv:2111.00396 [cs.LG], 2022.
    \bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.
    \bibitem{krizhevsky2012imagenet} - ImageNet Classification with Deep Convolutional Neural Networks
    \bibitem{simonyan2014very} - Very Deep Convolutional Networks for Large-Scale Image Recognition
    \bibitem{szegedy2015going} - Going Deeper with Convolutions
    \bibitem{he2016deep} - Deep Residual Learning for Image Recognition
    \bibitem{howard2017mobilenets} - MobileNets: Efficient Convolutional Neural Networks for Mobile Vision Applications
    \bibitem{tan2019efficientnet} - EfficientNet: Rethinking Model Scaling for Convolutional Neural Networks
    \bibitem{radosavovic2020designing} - Designing Network Design Spaces
    \bibitem{xie2021segformer} - SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers
    \bibitem{choromanski2020rethinking} - Rethinking Attention with Performers
    \bibitem{cheng2017survey} - A Survey of Model Compression and Acceleration for Deep Neural Networks
    \bibitem{hinton2015distilling} - Distilling the Knowledge in a Neural Network
    \bibitem{gu2022efficiently} A. Gu, K. Goel, and C. Ré, "Efficiently Modeling Long Sequences with Structured State Spaces," arXiv preprint arXiv:2111.00396, 2022.
    \bibitem{gu2020hippo} A. Gu, T. Dao, S. Ermon, A. Rudra, and C. Ré, "HiPPO: Recurrent Memory with Optimal Polynomial Projections," Advances in Neural Information Processing Systems, vol. 33, pp. 1474-1487, 2020.
    \bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with State Spaces," arXiv preprint arXiv:2302.01327, 2023.

@@

    \bibitem{bakas2019identifying} S. Bakas et al., "Identifying the Best Machine Learning Algorithms for Brain Tumor Segmentation, Progression Assessment, and Overall Survival Prediction in the BRATS Challenge," arXiv:1811.02629 [cs.CV], 2019.

@@

    \bibitem{DBLP:journals/corr/HeZRS15} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," CoRR, vol. abs/1512.03385, 2015. [Online]. Available: http://arxiv.org/abs/1512.03385

    \bibitem{zhang2022segvit} B. Zhang et al., "SegViT: Semantic Segmentation with Plain Vision Transformers," arXiv:2210.05844 [cs.CV], 2022.

    \bibitem{ultralightvmunet} R. Wu, Y. Liu, P. Liang, and Q. Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr. 20, 2024. [Online]. Available: https://arxiv.org/pdf/2403.20035.pdf
    \bibitem{zhao2017pyramid}
    H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
    \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
    \bibitem{xie2021segformer} E. Xie, W. Wang, Z. Yu, A. Anandkumar, J. Alvarez, and P. Luo, “SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers.” Available: https://arxiv.org/pdf/2105.15203.pdf
    \bibitem{krizhevsky2009learning} A. Krizhevsky, "Learning Multiple Layers of Features from Tiny Images," Tech Report, 2009.
    \bibitem{gu2023mamba} A. Gu and T. Dao, "Mamba: Linear-Time Sequence Modeling with Selective State Spaces," arXiv:2312.00752 [cs.LG], 2023.
    \bibitem{he2016deep} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," arXiv:1512.03385 [cs.CV], 2016.
    \bibitem{xie2017aggregated} S. Xie, R. Girshick, P. Dollár, Z. Tu, and K. He, "Aggregated Residual Transformations for Deep Neural Networks," arXiv:1611.05431 [cs.CV], 2017.
    \bibitem{krizhevsky2012imagenet} A. Krizhevsky, I. Sutskever, and G. E. Hinton, "ImageNet Classification with Deep Convolutional Neural Networks," Advances in Neural Information Processing Systems, vol. 25, 2012.
    \bibitem{sutskever2013importance} I. Sutskever, J. Martens, G. Dahl, and G. Hinton, "On the Importance of Initialization and Momentum in Deep Learning," Proceedings of the 30th International Conference on Machine Learning, vol. 28, pp. 1139-1147, 2013.
    \bibitem{lecundocument} Y. Lecun, L. Bottou, Y. Bengio and P. Haffner, "Gradient-based learning applied to document recognition," in Proceedings of the IEEE, vol. 86, no. 11, pp. 2278-2324, Nov. 1998, doi: 10.1109/5.726791.
    \bibitem{zhu2024vision} L. Zhu, B. Liao, Q. Zhang, X. Wang, W. Liu, and X. Wang, “Vision Mamba: Efficient Visual Representation Learning with Bidirectional State Space Model,” arXiv.org, Feb. 10, 2024. https://arxiv.org/abs/2401.09417   
\end{thebibliography}

>>>>

\documentclass[conference]{IEEEtran}
\IEEEoverridecommandlockouts
% The preceding line is only needed to identify funding in the first footnote. If that is unneeded, please comment it out.
\usepackage{cite}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{algorithmic}
\usepackage{graphicx}
\usepackage{textcomp}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{algorithm}
\usepackage{caption} % for caption setup

@@

\begin{abstract}
    This paper explores the performance of the Mamba architecture, a variant of Structured State Space Sequence Models (SSMs), in comparison to state-of-the-art models such as Transformers, U-Net, and ResNet for vision tasks. We focus on image segmentation and classification, with the goal of maintaining high accuracy while keeping parameter counts low. Developing compact models with fewer parameters is crucial for the future of AI, not only to reduce energy consumption but also to enable deployment on resource-constrained edge devices.
    By leveraging the selective state spaces in Mamba blocks, we aim to achieve efficient and effective segmentation and classification performance while maintaining the linear scalability of SSMs. We implement Mamba-based architectures and compare their results to popular models like U-Net, ResNet, and Transformer-based approaches on the tasks of segmentation and classification.
    This work highlights the potential of Mamba and SSMs for efficient vision tasks, contributing to the development of compact yet accurate models in image segmentation and classification. We open-source our code to facilitate further research and exploration of these architectures: \href{https://github.com/lxaw/mamba-vs-else-vision}{https://github.com/lxaw/mamba-vs-else-vision}
\end{abstract}

\begin{IEEEkeywords}
    Mamba, SSMs, Image Segmentation, Image Classification
\end{IEEEkeywords}

\section{Introduction}
The rapid advancement of deep learning has led to remarkable achievements in various domains, including computer vision, natural language processing, and speech recognition. However, this progress has been accompanied by a significant increase in the size and complexity of neural network models. State-of-the-art (SOTA) models often have several million or even billions of parameters, making them computationally expensive and challenging to deploy on resource-constrained devices \cite{modelSize}. This trend poses concerns regarding energy consumption and the feasibility of deploying these models on edge devices, which have limited memory and processing power.

@@

CNNs achieve their efficiency and effectiveness through their unique architecture and the use of convolutional operations. The convolutional layers in CNNs apply a set of learnable filters (kernels) to the input image or feature map. These filters slide over the input, performing element-wise multiplications and summing the results to produce a new feature map \cite{lecundocument}. This process allows CNNs to automatically learn and extract relevant features from the input data.

@@

The above derivation was done with reference to \cite{mambaNotes}.

@@

Now, the kernel can be constructed from the convolutional form \eqref{eq:convolutionalform} and output $y$ can be computed.
\begin{align}
    \mbfb{K} & = (\mathbf{C}\mbfb{B}, \mathbf{C}\mbfb{A}\mbfb{B}, \ldots, \mathbf{C}\mbfb{A}^k\mbfb{B}, \ldots) \\
    y        & = x * \mbfb{K}
\end{align}
The above derivation was done with reference to \cite{mambaNotes}.
This convolution can be computed in parallel as $y_k$ does not depend on previous outputs.

@@

For this our analysis, we compare four models against our own developed ones. The models are U-Net\cite{ronneberger2015unet}, ResNet18 \cite{he2016deep}, ResNet50, U-Mamba\cite{U-Mamba}, and SegFormer\cite{xie2021segformer}. Below we explain our choices.
\subsection{U-Net}
The core idea behind U-Net is to complement a traditional contracting network with successive layers that replace pooling operations with upsampling operators. These upsampling layers aim to increase the resolution of the output. Subsequent convolutional layers can then learn to assemble a precise output based on this high-resolution information \cite{ronneberger2015unet}.
A key modification in the U-Net architecture is the inclusion of a large number of feature channels in the upsampling part. This allows the network to effectively propagate contextual information to higher resolution layers. As a result, the expansive path of the network becomes more or less symmetric to the contracting part, leading to a distinctive U-shaped architecture. To predict pixels at the border regions of the image, the missing context is extrapolated by mirroring the input image. This tiling strategy is crucial for applying the network to large images, as it circumvents resolution limitations imposed by GPU memory constraints \cite{ronneberger2015unet}. Due to its encoder-decoder architecture, skip connections, multi-scale feature extraction, and its efficiency, U-Net has been used in many denoising and diffusion models. For instance, DDPMs (Denoising Diffusion Probabilistic Models) use an architecture similar to U-Net for denoising and sample generation \cite{DBLP:journals/corr/abs-2006-11239} while the popular Stable Diffusion architecture uses a U-Net based architecture for the diffusion process \cite{rombach2021highresolution}. We have chosen to compare our novel segmentation architecture against the U-Net architecture due to its well-established reputation and widespread adoption in the field of medical image segmentation.

\subsection{ResNet}
We have also selected ResNet18 and ResNet50 as additional benchmarks. These architectures, introduced by He et al. in their seminal work "Deep Residual Learning for Image Recognition" \cite{he2016deep}, have revolutionized the field of deep learning by addressing the problem of vanishing gradients in deep neural networks. The key innovation in ResNets is the introduction of residual connections, which allow the network to learn residual functions with reference to the input layer, thereby facilitating the training of much deeper networks. ResNet18 and ResNet50, with 18 and 50 layers respectively, have been widely adopted in various computer vision tasks, including image classification, object detection, and segmentation. These models have demonstrated exceptional performance and generalization ability across diverse datasets. By comparing our proposed architecture against ResNet18 and ResNet50, we aim to assess its effectiveness in relation to these well-established and highly influential architectures. This comparison will provide valuable insights into the capabilities of our model and its potential to advance the state-of-the-art in image segmentation tasks.
\subsection{U-Mamba}
As our model takes much inspiration from the recently developed U-Mamba design \cite{U-Mamba}, we also choose to incorporate it in our comparison. U-Mamba addresses these limitations by introducing a novel hybrid CNN-SSM block that leverages the strengths of both architectures. The convolutional layers in the block are responsible for local feature extraction, while the State Space Sequence Models (SSMs) \cite{gu2022efficiently}, a new family of deep sequence models, are known for their strong capability in handling long sequences and capturing long-range dependencies. By integrating these two components, U-Mamba achieves a balance between local and global information processing, enabling it to effectively handle long-range dependencies in biomedical image segmentation tasks. Moreover, U-Mamba incorporates a self-configuring mechanism that allows it to automatically adapt to various datasets without manual intervention, enhancing its versatility and usability.
\subsection{SegFormer}
SegFormer \cite{xie2021segformer} is a transformer-based semantic segmentation model that employs a hierarchical structure and a novel attention mechanism called Efficient Self-Attention (ESA). The SegFormer architecture consists of a transformer encoder for capturing long-range dependencies and a lightweight All-MLP decoder for generating high-resolution segmentation masks. The ESA module reduces the computational complexity of self-attention by performing attention operations in a local window and aggregating global information through a depth-wise convolution. This allows SegFormer to efficiently process high-resolution images while capturing both local and global context. Moreover, SegFormer introduces a position-sensitive embedding scheme that encodes both spatial and channel-wise information, enhancing the model's ability to capture fine-grained details. By comparing U-Mamba against SegFormer, we aim to evaluate the effectiveness of our hybrid CNN-SSM approach in relation to the transformer-based segmentation mechanism employed by SegFormer. This comparison will provide insights into the strengths and weaknesses of both architectures and their ability to handle complex biomedical image segmentation tasks. Furthermore, it will help us understand the potential of transformer-based approaches and hybrid approaches in advancing the state-of-the-art in biomedical image segmentation.

@@

\section{Results for Segmentation Task}
We now compare our models against ResNet18, ResNet50, standard U-Mamba bottleneck, UNet, and UltraLight VM-UNet. We train for 100 epochs on all models, and use dice-loss with Adam optimizer. We train on roughly 3000 images, and validate on roughly 400. we then test on roughly 400. We show the distributions of our data in Figure \ref{fig:datastats}.
Model results are shown in Table \ref{tab:model_comparison} and Figures \ref{fig:balls},\ref{fig:segs}.

\renewcommand{\arraystretch}{1.25}
\begin{table}[ht]
    \centering
    \captionsetup{skip=6pt}
    \begin{tabular}{|l|c|c|}
        \hline
        \textbf{Model Name} & \textbf{\#Params (M) $\downarrow$} & \textbf{Dice Score $\uparrow$} \\
        \hline
        UMambaBot-PP(Ours) & 10.0 & 0.8867 \\
        UMambaBot & 9.8 & 0.8799 \\
        ResNet18 & 13.9 & 0.8648 \\
        ResNet50 & 42.9 & 0.8672 \\
        SegFormer & 17.8 & 0.8454 \\
        UNet & \textbf{7.8} & \textbf{0.8929} \\
        \hline\hline
        UL-VM-UNet & 0.049 & 0.79955 \\
        UL-VM-UNet-v1(Ours) & 0.20 & 0.8409 \\
        UL-VM-UNet-v2(Ours) & \textbf{0.042} & 0.8243 \\
        UL-VM-UNet-v3(Ours) & 0.176 & 0.8109 \\
        UL-VM-UNet-v4(Ours) & 1.1 & 0.8548 \\
        UL-VM-UNet-v5(Ours) & 0.7 & \textbf{0.8633} \\
        UL-VM-UNet-v6(Ours) & 0.77 & 0.8628 \\
        \hline
    \end{tabular}
    \caption{Comparison of different segmentation models.}
    \label{tab:model_comparison}
\end{table}

@@

\section{Results for Classification Task}
To further demonstrate the viability and versatility of the Mamba architecture, we conducted experiments on the CIFAR-10 dataset \cite{krizhevsky2009learning}, a widely-used benchmark for image classification tasks. CIFAR-10 consists of 60,000 32x32 color images across 10 classes, with 6,000 images per class. The dataset is split into 50,000 training images and 10,000 test images.
For this experiment, we employed the Vision Mamba (ViM) architecture proposed by Zhu et al. \cite{zhu2024vision}. ViM is an efficient visual representation learning framework that leverages a bidirectional state space model to capture long-range dependencies in visual data. By combining the strengths of CNNs and SSMs, ViM aims to achieve competitive performance while maintaining computational efficiency.

@@

We compared the performance of ViM against the Vision Transformer (ViT) architecture \cite{dosovitskiy2021image} on the CIFAR-10 dataset. The ViM model had 1,974,664 parameters, while the ViT had 1,767,602.
Both models were trained for 100 epochs using the same experimental setup, including hyperparameters, optimization algorithm, and data augmentation techniques. Training and validation was done with batch size of 64, and upon every training iteration we do a validation iteration to obtain the validation scores. To see more on the training, please check our GitHub. The experiments were conducted on an NVIDIA GeForce RTX 3090 Ti GPU to ensure a fair comparison of training and validation times.
Figure \ref{fig:ViMViTF1} presents the validation F1 score and loss curves for both ViM and ViT over the course of 100 epochs, Figure \ref{fig:ViMViTLoss} shows the validation loss values over the 100 epochs, while Figure \ref{fig:ViMViTTrainingValidationTime} compares the training and validation times for both models. Note that ViM achieves similar performance with less training and validation time.

@@

\section{Conclusion}
We investigate the Mamba architecture with regards to the vision tasks of segmentation and classification. We develop and fine-tune models for both tasks, showing that they can outperform the SOTA in parameter count or accuracy. We hope to show that there are viable alternatives to transformers, and that 
transformer-level performance can be achieved with Mamba in the realm of computer vision.

\begin{thebibliography}{00}
    \bibitem{brown2020language} T. B. Brown et al., "Language Models are Few-Shot Learners," arXiv:2005.14165 [cs.CL], 2020.
    \bibitem{dosovitskiy2021image} A. Dosovitskiy et al., "An Image is Worth 16x16 Words: Transformers for Image Recognition at Scale," arXiv:2010.11929 [cs.CV], 2021.
    \bibitem{ronneberger2015unet} O. Ronneberger, P. Fischer, and T. Brox, "U-Net: Convolutional Networks for Biomedical Image Segmentation," arXiv:1505.04597 [cs.CV], 2015.
    \bibitem{he2016deep} K. He, X. Zhang, S. Ren, and J. Sun, "Deep Residual Learning for Image Recognition," arXiv:1512.03385 [cs.CV], 2016.
    \bibitem{vaswani2017attention} A. Vaswani et al., "Attention Is All You Need," arXiv:1706.03762 [cs.CL], 2017.
    \bibitem{simonyan2014very} K. Simonyan and A. Zisserman, “Published as a conference paper at ICLR 2015 VERY DEEP CONVOLUTIONAL NETWORKS FOR LARGE-SCALE IMAGE RECOGNITION.” Available: https://arxiv.org/pdf/1409.1556
    \bibitem{szegedy2015going} C. Szegedy et al., “Going deeper with convolutions.” Available: https://arxiv.org/pdf/1409.4842
    \bibitem{howard2017mobilenets} A. Howard et al., “MobileNets: Efficient Convolutional Neural Networks for Mobile Vision Applications.” Available: https://arxiv.org/pdf/1704.04861
    \bibitem{tan2019efficientnet} M. Tan and Q. Le, “EfficientNet: Rethinking Model Scaling for Convolutional Neural Networks.” Available: https://arxiv.org/pdf/1905.11946
    \bibitem{radosavovic2020designing} I. Radosavovic, R. Kosaraju, R. Girshick, K. He, and P. Dollár, “Designing Network Design Spaces.” Accessed: Apr. 30, 2024. [Online]. Available: https://arxiv.org/pdf/2003.13678
    \bibitem{choromanski2020rethinking} K. Choromanski et al., “Published as a conference paper at ICLR 2021 RETHINKING ATTENTION WITH PERFORMERS.” Accessed: Apr. 30, 2024. [Online]. Available: https://arxiv.org/pdf/2009.14794
    \bibitem{cheng2017survey} Y. Cheng, D. Wang, P. Zhou, and T. Zhang, “IEEE SIGNAL PROCESSING MAGAZINE, SPECIAL ISSUE ON DEEP LEARNING FOR IMAGE UNDERSTANDING 1 A Survey of Model Compression and Acceleration for Deep Neural Networks.” Accessed: Apr. 30, 2024. [Online]. Available: https://arxiv.org/pdf/1710.09282
    \bibitem{hinton2015distilling} G. Hinton, O. Vinyals, and J. Dean, “Distilling the Knowledge in a Neural Network,” 2015. Accessed: Apr. 30, 2024. [Online]. Available: https://arxiv.org/pdf/1503.02531
    \bibitem{gu2020hippo} A. Gu, T. Dao, S. Ermon, A. Rudra, and C. Ré, "HiPPO: Recurrent Memory with Optimal Polynomial Projections," Advances in Neural Information Processing Systems, vol. 33, pp. 1474-1487, 2020.

@@

@@

    \bibitem{ultralightvmunet} R. Wu, Y. Liu, P. Liang, and Q. Chang, “UltraLight VM-UNet: Parallel Vision Mamba Significantly Reduces Parameters for Skin Lesion Segmentation.” Accessed: Apr. 20, 2024. [Online]. Available: https://arxiv.org/pdf/2403.20035.pdf
    \bibitem{zhao2017pyramid} H. Zhao, J. Shi, X. Qi, X. Wang and J. Jia, "Pyramid Scene Parsing Network," 2017 IEEE Conference on Computer Vision and Pattern Recognition (CVPR), 2017, pp. 6230-6239, doi: 10.1109/CVPR.2017.660.
    \bibitem{dataset} “Brain MRI segmentation,” www.kaggle.com. https://www.kaggle.com/datasets/mateuszbuda/lgg-mri-segmentation
    \bibitem{xie2021segformer} E. Xie, W. Wang, Z. Yu, A. Anandkumar, J. Alvarez, and P. Luo, “SegFormer: Simple and Efficient Design for Semantic Segmentation with Transformers.” Available: https://arxiv.org/pdf/2105.15203.pdf
    \bibitem{krizhevsky2009learning} A. Krizhevsky, "Learning Multiple Layers of Features from Tiny Images," Tech Report, 2009.
    \bibitem{xie2017aggregated} S. Xie, R. Girshick, P. Dollár, Z. Tu, and K. He, "Aggregated Residual Transformations for Deep Neural Networks," arXiv:1611.05431 [cs.CV], 2017.
    \bibitem{krizhevsky2012imagenet} A. Krizhevsky, I. Sutskever, and G. E. Hinton, "ImageNet Classification with Deep Convolutional Neural Networks," Advances in Neural Information Processing Systems, vol. 25, 2012.
    \bibitem{lecundocument} Y. Lecun, L. Bottou, Y. Bengio and P. Haffner, "Gradient-based learning applied to document recognition," in Proceedings of the IEEE, vol. 86, no. 11, pp. 2278-2324, Nov. 1998, doi: 10.1109/5.726791.
    \bibitem{zhu2024vision} L. Zhu, B. Liao, Q. Zhang, X. Wang, W. Liu, and X. Wang, “Vision Mamba: Efficient Visual Representation Learning with Bidirectional State Space Model,” arXiv.org, Feb. 10, 2024. https://arxiv.org/abs/2401.09417   
    \bibitem{modelSize} H. Naveed et al., “A Comprehensive Overview of Large Language Models.” Available: https://arxiv.org/pdf/2307.06435
    \bibitem{mambaNotes} U. Jamil, “hkproj/mamba-notes,” GitHub, Apr. 27, 2024. https://github.com/hkproj/mamba-notes (accessed Apr. 30, 2024).
\end{thebibliography}
